---
title: "Loops and parallelisation"
---

```{r}
#| output: false

library(dplyr)
library(purrr)

set.seed(1234)
```

Loops let you repeat an operation many times — over the elements of a list, the rows of a dataset, or until some condition is met — without copy-pasting the same code over and over.

## Some data to work on

```{r}
patients <- tibble::tibble(
  id      = 1:8,
  age     = c(55, 62, 71, 48, 65, 70, 58, 44),
  score   = c(4.2, 5.1, 3.8, 6.3, 4.9, 3.5, 7.1, 5.5),
  group   = c("A","B","A","B","A","A","B","B")
)
patients
```

# For loops

"For each item in this sequence, do something":

```{r}
for (i in 1:5) {
  print(i^2)
}
```

A more practical use — printing a summary for each patient:

```{r}
for (i in 1:nrow(patients)) {
  cat("Patient", patients$id[i],
      "| age:", patients$age[i],
      "| score:", patients$score[i], "\n")
}
```

You can also loop over the values directly instead of using an index:

```{r}
for (g in unique(patients$group)) {
  n <- sum(patients$group == g)
  cat("Group", g, "has", n, "patients\n")
}
```

# While loops

Runs as long as a condition is `TRUE`. Useful when you don't know in advance how many iterations you'll need — but **be careful**: if the condition never becomes `FALSE`, you get an infinite loop.

```{r}
x <- 1
while (x < 100) {
  x <- x * 2
}
x
```

# The `apply()` family

R has built-in functions for applying a function across the elements of a list or the rows/columns of a matrix, without writing an explicit loop.

`sapply()` returns a simplified result (usually a vector):

```{r}
sapply(patients$score, function(x) round(x * 1.1, 2))
```

`lapply()` always returns a list:

```{r}
lapply(patients$score, function(x) round(x * 1.1, 2))
```

`apply()` works on matrices — `MARGIN = 1` for rows, `MARGIN = 2` for columns:

```{r}
m <- matrix(1:12, nrow = 3)
apply(m, MARGIN = 1, sum)  # row sums
apply(m, MARGIN = 2, sum)  # column sums
```

The downside is that different `apply` variants return different types (vector, list, matrix…) and have slightly different argument conventions, which can trip you up.

# `purrr::map()`

`purrr::map()` is consistent: **it always returns a list**. For different output types, there are type-safe variants:

```{r}
# Always a list:
map(patients$score, \(x) round(x * 1.1, 2))

# Always a numeric vector:
map_dbl(patients$score, \(x) round(x * 1.1, 2))

# Always a character vector:
map_chr(patients$score, \(x) paste0("score: ", x))
```

The pattern is always `map_<type>(list, function)`. You always know what type you'll get back — no surprises.

You can map over two inputs at once with `map2()`:

```{r}
map2_chr(
  patients$id, patients$score,
  \(id, sc) paste0("Patient ", id, ": ", sc)
)
```

Or over an arbitrary number of inputs with `pmap()` (pass a list of input vectors):

```{r}
pmap_chr(
  list(patients$id, patients$age, patients$score),
  \(id, age, sc) paste0("ID ", id, ", age ", age, ", score ", sc)
)
```

# `Vectorize()`

Last time we wrote our own functions. Some functions only handle one value at a time and break when given a vector:

```{r}
#| error: true

classify <- function(x) {
  if (x > 5) "high" else "normal"
}

classify(patients$score)  # Error — can't handle a vector
```

`Vectorize()` wraps the function so it works element-by-element across a whole vector:

```{r}
classify_vec <- Vectorize(classify)
classify_vec(patients$score)
```

Note: `map_chr()` or `ifelse()` can often do this more cleanly, but `Vectorize()` is a handy shortcut when you already have a function written and just want it to work on vectors.

# Parallel loops with `foreach`

When you have many iterations or slow computations, you can run them in parallel — across multiple CPU cores at once. The `{foreach}` and `{doParallel}` packages make this straightforward:

```{r}
#| eval: false

library(foreach)
library(doParallel)

cl <- makeCluster(2)       # spin up 2 CPU cores
registerDoParallel(cl)

results <- foreach(sc = patients$score) %dopar% {
  Sys.sleep(0.1)           # imagine this is slow
  sc * 1.1
}

stopCluster(cl)

results
```

The key idea: replace `%do%` (serial) with `%dopar%` (parallel) and the rest looks almost identical to a regular for loop. How much you gain depends on how many cores you have and how slow each iteration is.

# Honourable mention: recursion

A function can call *itself*. This is called **recursion**. It's elegant for certain problems — tree structures, divide-and-conquer algorithms, the Fibonacci sequence — but in everyday data analysis you'll rarely need it. It can also be slow and harder to debug than a plain loop.

```{r}
# Classic example: factorial
factorial_r <- function(n) {
  if (n <= 1) return(1)
  n * factorial_r(n - 1)
}

factorial_r(5)  # 5! = 120
```

# Further reading

[purrr cheatsheet](https://posit.co/resources/cheatsheets/){.external target="_blank"}

[R for Data Science — Iteration chapter](https://r4ds.hadley.nz/iteration.html){.external target="_blank"}

[foreach and parallel R (CRAN vignette)](https://cran.r-project.org/web/packages/foreach/vignettes/foreach.html){.external target="_blank"}
