---
title: "Loops and parallelisation"
---

```{r}
#| output: false

library(dplyr)
library(purrr)
set.seed(1234)
```

Loops are one of the most fundamental tools in programming.
They allow you to **repeat** a block of code — either a fixed number of times, or until some condition is met.

## Some data to work on
```{r}
patients <- tibble(
  id      = 1:8,
  age     = c(45, 67, 53, 71, 38, 60, 55, 49),
  systolic = c(120, 145, 132, 158, 112, 140, 128, 119)
)
patients
```

# `for` loops

A `for` loop runs a block of code once for every element in a vector:
```{r}
for (i in 1:5) {
  print(i)
}
```

You can loop over any vector — including character vectors:
```{r}
for (col in c("age", "systolic")) {
  cat(col, "— mean:", mean(patients[[col]]), "\n")
}
```

A common pattern is to build up a result vector as you go:
```{r}
high_bp <- c()
for (i in seq_len(nrow(patients))) {
  high_bp[i] <- patients$systolic[i] >= 130
}
high_bp
```

# `while` loops

A `while` loop keeps running as long as a condition is `TRUE`:
```{r}
x <- 2
while (x < 100) {
  x <- x^2
  print(x)
}
```

::: callout-warning
Be careful: if the condition never becomes `FALSE`, the loop runs forever.
:::

# The `apply()` family

`for` loops are clear but can be verbose. The `apply()` family lets you apply a function to every element (or row/column) of an object in one go.

## `sapply()` and `lapply()`

Both apply a function to each element of a vector or list. The difference is the output format:
```{r}
# lapply always returns a list
lapply(patients[c("age", "systolic")], mean)
```

```{r}
# sapply tries to simplify to a vector or matrix
sapply(patients[c("age", "systolic")], mean)
```

## `apply()` for matrices/data frames

`apply()` works on rows (`MARGIN = 1`) or columns (`MARGIN = 2`):
```{r}
mat <- matrix(1:12, nrow = 3)
apply(mat, MARGIN = 1, FUN = sum) # row sums
apply(mat, MARGIN = 2, FUN = sum) # col sums
```

The problem with the `apply()` family is that the function names and output types are different every time, which makes them hard to remember.

# `purrr::map()`

The `{purrr}` package gives you a consistent, predictable alternative.
`map()` always returns a list; suffix variants guarantee the output type:

| Function     | Output           |
|--------------|------------------|
| `map()`      | list             |
| `map_dbl()`  | numeric vector   |
| `map_chr()`  | character vector |
| `map_lgl()`  | logical vector   |
| `map_df()`   | data frame       |

```{r}
# Always returns a list
map(patients[c("age", "systolic")], mean)
```

```{r}
# Want a numeric vector? Be explicit:
map_dbl(patients[c("age", "systolic")], mean)
```

```{r}
# Works naturally with pipes and anonymous functions
patients[c("age", "systolic")] |>
  map_dbl(\(col) round(mean(col), 1))
```

## `map2()` — mapping over two inputs at once

```{r}
names  <- c("Alice", "Bob", "Carol")
scores <- c(88, 95, 72)

map2_chr(names, scores, \(n, s) paste0(n, " scored ", s))
```

# `Vectorize()`

Last session we wrote functions. Most built-in R functions are already *vectorised* — they work on a whole vector at once:
```{r}
sqrt(c(4, 9, 16))
```

But if you write your own function with an `if` / `else` inside, it usually expects a single value:
```{r}
#| error: true

bp_category <- function(systolic) {
  if (systolic >= 130) "High" else "Normal"
}

bp_category(c(120, 145, 132)) # doesn't work on a vector
```

`Vectorize()` wraps your function so it loops automatically:
```{r}
bp_category_v <- Vectorize(bp_category)
bp_category_v(patients$systolic)
```

This is equivalent to writing a `for` loop, but in one line. (For the tidyverse crowd, `dplyr::case_when()` is often a cleaner solution inside `mutate()`.)

# Parallel loops with `foreach` and `%dopar%`

So far, all loops run **sequentially** — one step at a time.
For computationally expensive work (e.g., bootstrapping, simulations), you can run iterations **in parallel** across multiple CPU cores.

```{r}
#| eval: false

library(foreach)
library(doParallel)

cl <- makeCluster(2)          # open 2 workers
registerDoParallel(cl)

results <- foreach(i = 1:8, .combine = c) %dopar% {
  patients$systolic[i]^2
}

stopCluster(cl)
closeAllConnections()

results
```

The key difference is `%dopar%` instead of `%do%`: the former distributes iterations across cores, the latter runs them sequentially (useful for testing before going parallel).

::: callout-tip
For simpler parallelism inside `purrr`, the [`{furrr}`](https://furrr.futureverse.org/) package lets you swap `map()` for `future_map()` with almost no code change.
:::

# Honourable mention: Recursion

A function can call *itself*. This is called **recursion**, and it's another way to repeat an operation. Classic example: factorials.

```{r}
factorial_r <- function(n) {
  if (n <= 1) return(1)
  n * factorial_r(n - 1)
}
factorial_r(5)
```

Recursion is elegant for certain problems (tree traversal, divide-and-conquer algorithms) but can be slow and hard to debug in R. Loops or `map()` are usually the better choice for day-to-day data work.

# Further reading

[R for Data Science — Iteration](https://r4ds.hadley.nz/iteration.html){.external target="_blank"}

[purrr documentation](https://purrr.tidyverse.org/){.external target="_blank"}

[foreach + doParallel vignette](https://cran.r-project.org/web/packages/foreach/vignettes/foreach.html){.external target="_blank"}
