---
title: "Loops and parallelisation"
---

```{r}
#| output: false

library(dplyr)
library(purrr)
set.seed(1234)
```

Loops are one of the most fundamental tools in programming.
They allow you to **repeat** a block of code — either a fixed number of times, or until some condition is met.

## Some data to work on
```{r}
patients <- tibble(
  id      = 1:8,
  age     = c(45, 67, 53, 71, 38, 60, 55, 49),
  systolic = c(120, 145, 132, 158, 112, 140, 128, 119)
)
patients |> gt::gt(id = "patients")
```

# `for` loops

A `for` loop runs a block of code once for every element in a vector; i.e., it runs a fixed/prespecified number of times:
```{r}
for (i in 1:5) {
  print(
    paste0(
      "Age of row ", i, ": ", 
      patients$age[i]
    )
  )
}
```

You can loop over any vector — including character vectors:
```{r}
columns <- c("age", "systolic")
for (col in columns) {
  print(paste0(
    "Mean ", col, ": ",
    mean(patients[[col]]))
  )
}
```

A common pattern is to build up a result vector as you go:
```{r}
high_bp <- c()

for (i in 1:nrow(patients)) {
  high_bp <- c(high_bp, patients$age[i])
}

high_bp
```

This might be useful if you're bootstrapping, for instance:
```{r}
#| eval: false

results <- c()

mean_age <- function(df) mean(df$age)

for (i in 1:10000) {
  dataset_i <- slice_sample(df, n = nrow(df), replace = TRUE) 
  result_i <- mean_age(dataset_i) # or any prespecified analysis function
  results <- c(results, result_i)
}
conf.int <- quantile(results, probs = c(0.025,0.975))

paste0("Mean age: ", mean(df$age))
conf.int

```

# `while` loops

A `while` loop keeps running as long as a condition (in this example `x < 1000`) is `TRUE`:
```{r}
x <- 2
while(x < 1000){
  x <- x^2
  print(x)
}
```

::: callout-warning
Be careful: if the condition never becomes `FALSE`, the loop runs forever.
:::

# The `apply()` family

`for` loops are clear but can be verbose. The `apply()` family lets you apply a function to every element (or row/column) of an object in one go.

## `sapply()` and `lapply()`

Both apply a function to each element of a vector or list. The difference is the output format.

`lapply()` always returns a list
```{r}
df <- patients |> select(-id)

lapply(X = df, FUN = mean)
```

`sapply()` tries to simplify to a vector or matrix
```{r}
sapply(X = df, FUN = mean)
```

This is the same as manually doing `lapply` + `unlist`:
```{r}
lapply(X = df, FUN = mean) |> unlist()
```

The `apply()` family also easily work with custom functions:
```{r}
squared <- function(x) {
  x^2
}
sapply(X = c(1:5), FUN = squared)
```

## `apply()` for matrices/data frames

`apply()` works on rows (`MARGIN = 1`) or columns (`MARGIN = 2`):
```{r}
mat <- matrix(1:12, nrow = 3)
mat

print("Row sums:")
apply(X = mat, MARGIN = 1, FUN = sum)

print("Column sums:")
apply(X = mat, MARGIN = 2, FUN = sum)
```

The problem with the `apply()` family is that the function names and output types are different every time, which makes them hard to remember.

# `purrr::map()`

The `{purrr}` package gives you a consistent, predictable alternative.
`map()` always returns a list; suffix variants guarantee the output type:

| Function     | Output           |
|--------------|------------------|
| `map()`      | list             |
| `map_dbl()`  | numeric vector   |
| `map_chr()`  | character vector |
| `map_lgl()`  | logical vector   |
| `map_df()`   | data frame       |

`map()` always returns a list
```{r}
map(df, mean)
```

Want a numeric vector? Be explicit:
```{r}
map_dbl(df, mean)
```

`map()` works naturally with pipes
```{r}
df |> map(mean)
```

## `map2()` — mapping over two inputs at once

This function requires both inputs to be the same length, and element i of input .x will be mapped to element i of input .y; that is, you don't get every possible combination of .x and .y
```{r}
names <- c("Alice", "Bob", "Carol")
scores <- c(88, 95, 72)

myPaste <- function(nm, scr) paste0(nm, " scored ", scr)

map2_chr(.x = names, .y = scores, .f = myPaste)
```

For all combinations, try something like:
```{r}
with(
  expand.grid(names, scores),
  paste0(Var1, " scored ", Var2)
)
```


# `Vectorize()`

Last session we wrote functions. Most built-in R functions are already *vectorised* — they work on a whole vector at once:
```{r}
sqrt(c(4, 9, 16))
```

But if you write your own function with an `if` / `else` inside, it usually expects a single value:
```{r}
#| error: true

bp_category <- function(systolic) {
  if (systolic >= 130) "High" else "Normal"
}

c(120, 145, 132) |> bp_category() # doesn't work on a vector
```

`Vectorize()` wraps your function so it loops automatically:
```{r}
bp_category <- bp_category |> Vectorize()

c(120, 145, 132) |> bp_category()
```

This is equivalent to writing a `for` loop, but in one line. (For the tidyverse crowd, `dplyr::case_when()` is often a cleaner solution inside `mutate()`.)

# Parallel loops with `foreach` and `%dopar%`

So far, all loops run **sequentially** — one step at a time.
For computationally expensive work (e.g., bootstrapping, simulations), you can run iterations **in parallel** across multiple CPU cores using `{doParallel}` with the for-like syntax of `foreach()`.
```{r}
#| output: false
library(foreach)
library(doParallel)
```

```{r}
cl <- makeCluster(10)          # open 10 workers
registerDoParallel(cl)

mean_age <- function(df) mean(df$age)

results <- foreach(i = 1:1000, 
                   .packages = "dplyr",
                   .combine = c) %dopar% {
  dataset_i <- slice_sample(df, n = nrow(df), replace = TRUE) 
  result_i <- mean_age(dataset_i)
  return(result_i)
}

stopCluster(cl)
closeAllConnections()

results |> quantile(probs = c(0.025,0.975))
```

`.combine = c` means the result of each loop iteration is returned in a combined vector (`c()`) in the end.

::: callout-tip
For simpler parallelism inside `purrr`, the [`{furrr}`](https://furrr.futureverse.org/) package lets you swap `map()` for `future_map()` with almost no code change.
:::

# Honourable mention: Recursion

A function can call *itself*. This is called **recursion**, and it's another way to repeat an operation. Classic example: factorials.

```{r}
factorial <- function(n) {
  if (n <= 1) return(1)
  else return(n * factorial(n - 1))
}
factorial(5)
```

Recursion is elegant for certain problems (tree traversal, divide-and-conquer algorithms) but can be slow and hard to debug in R. Loops or `map()` are almost always what you need.

# Further reading

[R for Data Science — Iteration](https://r4ds.hadley.nz/iteration.html){.external target="_blank"}

[purrr documentation](https://purrr.tidyverse.org/){.external target="_blank"}

[foreach + doParallel vignette](https://cran.r-project.org/web/packages/foreach/vignettes/foreach.html){.external target="_blank"}
