---
title: "Loops, parallelisation and recursion"
---

```{r}
#| output: false

library(dplyr)
library(purrr)
set.seed(1234)
```

Loops let you repeat an operation many times without copy-pasting code. They are the backbone of most data-processing pipelines.

## Some data to work on

A small dataset of clinical measurements, one row per patient:

```{r}
patients <- tibble::tibble(
  id      = 1:8,
  age     = c(45, 62, 38, 71, 55, 49, 66, 34),
  sbp     = c(128, 145, 118, 160, 135, 122, 150, 112),
  dbp     = c(82,  95,  75,  98,  88,  79,  92,  70)
)
patients
```

# `for` loops

A `for` loop iterates over a sequence and runs the same code block each time:

```{r}
for (i in 1:5) {
  print(i^2)
}
```

A practical example—flag patients with high systolic blood pressure (SBP ≥ 140):

```{r}
patients$hypertensive <- FALSE

for (i in seq_len(nrow(patients))) {
  if (patients$sbp[i] >= 140) {
    patients$hypertensive[i] <- TRUE
  }
}
patients
```

Note that this particular example could be done more neatly with `dplyr::mutate()`, but the loop makes the logic very transparent.

# `while` loops

A `while` loop keeps running as long as a condition is `TRUE`. Useful when you don't know in advance how many iterations you need:

```{r}
dose <- 10        # starting dose (mg)
effect <- 0

while (effect < 80) {
  dose   <- dose * 1.5
  effect <- 100 * (1 - exp(-dose / 30))   # simple exponential saturation curve
}
cat("Dose needed:", round(dose, 1), "mg  |  Effect:", round(effect, 1), "%\n")
```

::: {.callout-warning}
A `while` loop that never reaches a stopping condition will run forever. Always make sure the condition *can* eventually become `FALSE`.
:::

# The `apply()` family

`apply()`, `sapply()`, and `lapply()` are R's built-in way to apply a function to every element of a vector or list—replacing many `for` loops with a single line.

## `sapply()` — simplest to read

Returns a vector (or matrix) when the result has a regular shape:

```{r}
# Mean of each numeric column
sapply(patients[, c("age", "sbp", "dbp")], mean)
```

## `lapply()` — always returns a list

Useful when the results are of different lengths or types:

```{r}
lapply(patients[, c("age", "sbp", "dbp")], summary)
```

## `apply()` — over rows or columns of a matrix/data.frame

`MARGIN = 1` → rows, `MARGIN = 2` → columns:

```{r}
# Range (max – min) of each column
apply(patients[, c("age", "sbp", "dbp")], MARGIN = 2, FUN = function(x) diff(range(x)))
```

One pain point: the output type of `sapply()` is unpredictable—it silently returns a vector, a matrix, or a list depending on the result. This can cause hard-to-find bugs.

# `purrr::map()`

`{purrr}` provides a consistent, predictable alternative. The key rule: **the suffix tells you the output type**:

| function      | output        |
|---------------|---------------|
| `map()`       | always a list |
| `map_dbl()`   | numeric vector |
| `map_int()`   | integer vector |
| `map_chr()`   | character vector |
| `map_lgl()`   | logical vector |
| `map_dfr()`   | data frame (row-bind) |

```{r}
# Always returns a list — equivalent to lapply()
map(patients[, c("age", "sbp", "dbp")], mean)
```

```{r}
# Always returns a numeric vector — no surprises
map_dbl(patients[, c("age", "sbp", "dbp")], mean)
```

```{r}
# Two inputs at once: map2()
map2_dbl(patients$sbp, patients$dbp,
         \(s, d) s - d)   # pulse pressure
```

The predictability of `map_*()` is its biggest advantage over `sapply()`: you know exactly what type you'll get back, and if the result doesn't match, you get a clear error instead of a silent type change.

# `Vectorize()`

Last time we wrote functions that operate on a single value. But what if you want to pass a whole vector and get a vector back, without writing a loop yourself?

```{r}
# A function that only works on scalars
bp_category <- function(sbp) {
  if      (sbp < 120) "Normal"
  else if (sbp < 130) "Elevated"
  else if (sbp < 140) "Stage 1"
  else                "Stage 2"
}
```

Calling it on a vector fails:

```{r}
#| error: true
bp_category(patients$sbp)
```

`Vectorize()` wraps the function so that it loops over each element automatically:

```{r}
bp_category_v <- Vectorize(bp_category)
bp_category_v(patients$sbp)
```

This is handy for quick fixes, though `dplyr::case_when()` or `ifelse()` chains are usually cleaner for data-frame work.

# Parallel loops with `{foreach}` and `{doParallel}`

When a loop is slow because it's doing a lot of independent work, you can run the iterations simultaneously on multiple CPU cores.

```{r}
#| eval: false

library(doParallel)

cl <- makeCluster(2)          # open 2 worker processes
registerDoParallel(cl)

results <- foreach(x = patients$sbp, .combine = c) %dopar% {
  # each iteration runs on a separate core
  x^2
}

stopCluster(cl)
closeAllConnections()

results
```

The `%dopar%` operator swaps in parallel execution; swap back to `%do%` for sequential (useful for debugging). Note that starting workers has overhead, so parallelisation only pays off for genuinely slow tasks.

# Honourable mention: recursion

A recursive function calls *itself*. This can be a very elegant way to solve problems that have a naturally self-similar structure (e.g., navigating a folder tree, parsing nested data).

```{r}
factorial_r <- function(n) {
  if (n <= 1) return(1)
  n * factorial_r(n - 1)
}
factorial_r(6)   # 720
```

In practice, R's function-call overhead means recursion is often slower than a loop, and R has a relatively shallow default call stack (try `factorial_r(10000)`). For everyday data analysis, loops or `map()` are almost always a better choice.

# Further reading

[R for Data Science — Iteration chapter](https://r4ds.hadley.nz/iteration.html){.external target="_blank"}

[`{purrr}` reference](https://purrr.tidyverse.org/){.external target="_blank"}

[`{foreach}` vignette](https://cran.r-project.org/web/packages/foreach/vignettes/foreach.html){.external target="_blank"}
