---
title: "Coding paradigms"
---

```{r}
#| output: false

library(dplyr)
library(data.table)
set.seed(12345)
```

A **coding paradigm** is a style of coding that follows a specific set of principles; think of them like different dialects of the language. R has three common ones for data manipulation: 

| Paradigm       | Main package(s)     | Tagline                         |
|----------------|---------------------|---------------------------------|
| **Base R**     | Built-in            | Explicit, no dependencies       |
| **Tidyverse**  | `{dplyr}`, `{tidyr}` | Readable, pipe-friendly         |
| **data.table** | `{data.table}`      | Concise, fast, memory-efficient |

What's "best" depends on the use case. You'll encounter all three in the wild, although many guides use the Tidyverse.

## Some data to work on

```{r}
df <- tibble(
  id = 1:10,
  age = rnorm(10, mean = 40, sd = 10) |> round(),
  sex = sample(c("M", "F"), 10, replace = TRUE),
  weight = rnorm(10, mean = 75, sd = 10) |> round(),
  height = rnorm(10, mean = 1.66, sd = 0.1) |> round(2)
)

DT <- as.data.table(df)   # data.table copy

df |> gt::gt(id = "df_table")
```

# Indexing styles

### Base R

Base R uses explicit indexing with `$` for columns and `[]` for rows and columns. You need to specify exactly what you want, which can be verbose. `data.frame[rows, columns]` is the general form.

```{r}
df[df$age <= 35, "weight"]
df[df$age <= 35,]$weight
df[df$age <= 35,][["weight"]]
```

### Tidyverse

The tidyverse uses something called non-standard evaluation, allowing you to write column names directly without quoting. The pipe operator `|>` (or previously* `%>%`) lets you chain operations together in a readable way. You subset data using `filter()` (rows) and `select()` (columns).

```{r}
df |> filter(age <= 35) |> select(weight)
```

### Data.table

`{data.table}` extends the `[]` operator with a concise syntax: `DT[i, j, by]`, where 

- `i` filters rows, 
- `j` specifies what to do (select or compute columns), and 
- `by` groups the data. 

For example:

```{r}
DT[age <= 35, weight]
```

# Per-group operations

### Base R

Base R doesn't have built-in group-wise operations, but you can use `aggregate()` to summarise by group:
```{r}
aggregate(weight ~ sex, data = df, FUN = mean)
```

### Tidyverse

In the tidyverse, you can easily group by a variable and summarise:

```{r}
df |>
  group_by(sex) |>
  summarise(mean_weight = mean(weight))
```

### Data.table

In data.table, you can specify the grouping variable very naturally in the `by` argument:
```{r}
DT[, .(mean_weight = mean(weight)), by = sex]
```

in `{data.table}`, `.()` is shorthand for `list()` and names the output columns.

# Add a column

### Base R

In base R, you can add a new column by direct assignment:
```{r}
df$bmi <- (df$weight / df$height^2) |> round(1)
head(df)
``` 

### Tidyverse

In the tidyverse, you can use `mutate()` to add a new column:
```{r}
df <- df |> mutate(
  bmi = (weight / height^2) |> round(1)
)
head(df)
```

### Data.table

In data.table, you can add a new column using `:=`, which modifies the table in place:

```{r}
DT[, bmi := (weight / height^2) |> round(1)]
head(DT)
```

`:=` modifies the table **in place** (no copy is made), which is why data.table is so memory-efficient.


# When to use which?

## Base R

Base R uses **explicit indexing** and **standard evaluation**. What you see is what you get — there are no hidden conventions or non-standard evaluation tricks.

**Pros**: Zero dependencies; great for simple scripts, package development, and teaching fundamentals.  
**Cons**: Can get messy for complex data manipulation; less efficient on large datasets due to copying.

## Tidyverse 

The tidyverse (centred on `{dplyr}`) is **declarative** and uses **non-standard evaluation** — you write column names without quoting them or prefixing with `$`. Operations are chained together with the pipe `|>`, reading top-to-bottom like a sentence.

The pipe chains verbs together. Each verb takes a data frame as its first argument and returns a data frame — making them easy to compose.


**Pros**: The most readable option; ideal where clarity matters. Rich ecosystem of packages for everything from data manipulation to visualization.  
**Cons**: Like base R, it can be slower than data.table for large datasets due to copying.

## data.table 

data.table is **concise** and **fast**. It uses a unique `DT[i, j, by]` syntax that allows you to filter rows, operate on columns, and group by variables all in one place. It modifies data in place, which is why it's so memory-efficient. It also uses non-standard evaluation, so you can write column names directly without quoting.

**Pros**: Best when working with millions of rows; its in-place modification and optimised grouping make it the fastest and most memory-efficient option for large datasets. It also has powerful features for complex data manipulation.  
**Cons**: The concise syntax can be less intuitive for beginners, and it has a steeper learning curve. Speed gains are often exaggerated, as base R and tidyverse have gained performance improvements in recent years.

##
::: callout-tip
Most real-world R code mixes paradigms. You might load data with base R, clean it with `{dplyr}`, and run a heavy aggregation with `{data.table}`. Pick the right tool for each job.
:::

# Further reading

[R intro (CRAN)](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf){.external target="_blank"}

[Tidyverse paper](https://cran.r-project.org/web/packages/tidyverse/vignettes/paper.html){.external target="_blank"}

[dplyr intro vignette](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html){.external target="_blank"}

[data.table intro vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html){.external target="_blank"}
