---
title: "Coding paradigms"
---

```{r}
#| output: false

library(dplyr)
library(data.table)
set.seed(1234)
```

A **coding paradigm** is a style or philosophy for writing code — a set of conventions and patterns that shape how you think about and solve problems.

In R, there are three main paradigms for data manipulation:

| Paradigm       | Main package(s)     | Tagline                         |
|----------------|---------------------|---------------------------------|
| **Base R**     | Built-in            | Explicit, no dependencies       |
| **Tidyverse**  | `{dplyr}`, `{tidyr}` | Readable, pipe-friendly         |
| **data.table** | `{data.table}`      | Concise, fast, memory-efficient |

None is objectively best. Each has strengths, and you will encounter all three in the wild.

## Some data to work on

```{r}
df <- tibble(
  id      = 1:10,
  age     = c(15, 34, 22, 45, 17, 60, 29, 38, 16, 51),
  sex     = c("M","F","M","F","M","F","M","F","M","F"),
  weight  = c(55, 68, 80, 72, 60, 65, 90, 75, 58, 70),
  height  = c(1.65, 1.70, 1.80, 1.65, 1.72, 1.60, 1.85, 1.68, 1.69, 1.63)
)

DT <- as.data.table(df)   # data.table copy

df |> gt::gt(id = "df_table")
```

# Base R

Base R uses **explicit indexing** and **standard evaluation**. What you see is what you get — there are no hidden conventions or non-standard evaluation tricks.

## Filter rows

```{r}
df[df$age >= 18, ]
```

The logical vector `df$age >= 18` selects only the matching rows. The trailing `,` keeps all columns.

## Add a new column

```{r}
df$bmi <- df$weight / df$height^2
head(df)
```

Direct assignment to `df$newcol` is the base R way.

## Summarise by group

```{r}
aggregate(bmi ~ sex, data = df, FUN = mean)
```

`aggregate()` takes a formula, a data frame, and a summary function.

# Tidyverse

The tidyverse (centred on `{dplyr}`) is **declarative** and uses **non-standard evaluation** — you write column names without quoting them or prefixing with `$`. Operations are chained together with the pipe `|>`, reading top-to-bottom like a sentence.

## Filter rows

```{r}
df |> filter(age >= 18)
```

## Add a new column

```{r}
df <- df |> mutate(
  bmi = weight / height^2
)
head(df)
```

## Summarise by group

```{r}
df |>
  group_by(sex) |>
  summarise(mean_bmi = mean(bmi))
```

The pipe chains verbs together. Each verb takes a data frame as its first argument and returns a data frame — making them easy to compose.

# data.table

`{data.table}` extends the standard `[` operator with the concise syntax `DT[i, j, by]`:

- **i** — which rows (filter)
- **j** — what to do (select / compute columns)
- **by** — grouping variable

This makes it extremely fast and memory-efficient — a key advantage for large datasets.

## Filter rows

```{r}
DT[age >= 18]
```

## Add a new column

```{r}
DT[, bmi := weight / height^2]
head(DT)
```

`:=` modifies the table **in place** (no copy is made), which is why data.table is so memory-efficient.

## Summarise by group

```{r}
DT[, .(mean_bmi = mean(bmi)), by = sex]
```

`.()` is shorthand for `list()` and names the output columns.

# Side-by-side comparison

The same three operations, across all three paradigms:

```{r}
#| eval: false

## --- Filter: keep adults only ---

# Base R
df[df$age >= 18, ]

# Tidyverse
df |> filter(age >= 18)

# data.table
DT[age >= 18]


## --- New column: compute BMI ---

# Base R
df$bmi <- df$weight / df$height^2

# Tidyverse
df <- df |> mutate(bmi = weight / height^2)

# data.table
DT[, bmi := weight / height^2]


## --- Summarise: mean BMI by sex ---

# Base R
aggregate(bmi ~ sex, data = df, FUN = mean)

# Tidyverse
df |> group_by(sex) |> summarise(mean_bmi = mean(bmi))

# data.table
DT[, .(mean_bmi = mean(bmi)), by = sex]
```

# When to use which?

- **Base R** — zero dependencies; great for simple scripts, package development, and teaching fundamentals.
- **Tidyverse** — the most readable option; ideal for exploratory analysis and collaborative projects where clarity matters.
- **data.table** — best when working with millions of rows; its in-place modification and optimised grouping make it significantly faster.

::: callout-tip
Most real-world R code mixes paradigms. You might load data with base R, clean it with dplyr, and run a heavy aggregation with data.table. Pick the right tool for each job.
:::

# Further reading

[R intro (CRAN)](https://cran.r-project.org/doc/manuals/r-release/R-intro.pdf){.external target="_blank"}

[Tidyverse paper](https://cran.r-project.org/web/packages/tidyverse/vignettes/paper.html){.external target="_blank"}

[dplyr intro vignette](https://cran.r-project.org/web/packages/dplyr/vignettes/dplyr.html){.external target="_blank"}

[data.table intro vignette](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html){.external target="_blank"}
