---
title: "More Tidyverse"
---

```{r}
#| output: false

library(tidyverse)
library(palmerpenguins)
```

The **tidyverse** is a family of packages built to make data work in R readable, consistent, and pipe-friendly.

- `dplyr`: transform and summarise tables
- `tidyr`: reshape data between wide and long formats
- the rest: import, clean strings/factors, iterate, and make tidy tables

## `dplyr` in 2 minutes

`dplyr` gives you a small set of verbs that cover most day-to-day data wrangling:

- `select()` pick columns
- `filter()` pick rows
- `mutate()` create/change columns
- `summarise()` reduce to summaries (often after `group_by()`)
- `arrange()` sort rows

```{r}
penguins |>
  select(species, island, bill_length_mm, body_mass_g) |>
  filter(!is.na(body_mass_g)) |>
  group_by(species, island) |>
  summarise(mean_mass = round(mean(body_mass_g), 1), .groups = "drop") |>
  arrange(desc(mean_mass))
```

## A couple of neat `tidyr` functions

### `pivot_longer()` and `pivot_wider()`

Use these when your columns should become rows (or vice versa).

```{r}
wide_penguins <- penguins |>
  select(species, bill_length_mm, bill_depth_mm, flipper_length_mm) |>
  slice(1:5)

long_penguins <- wide_penguins |>
  pivot_longer(
    cols = c(bill_length_mm, bill_depth_mm, flipper_length_mm),
    names_to = "measure",
    values_to = "value"
  )

long_penguins
```

```{r}
long_penguins |>
  pivot_wider(names_from = measure, values_from = value)
```

### `separate_wider_delim()`

Great when one column stores multiple pieces of information.

```{r}
tibble(id = c("A-001", "B-014", "C-120")) |>
  separate_wider_delim(id, delim = "-", names = c("site", "participant"))
```

## Other tidyverse packages: purpose + useful functions

| Package | Purpose | Useful functions |
|---|---|---|
| `{readr}` | Fast reading/writing tabular files | `read_csv()`, `write_csv()`, `read_tsv()` |
| `{tibble}` | Modern data frame tools | `tibble()`, `as_tibble()`, `add_column()` |
| `{stringr}` | Consistent string handling | `str_detect()`, `str_replace()`, `str_to_lower()` |
| `{forcats}` | Factor cleaning/reordering | `fct_recode()`, `fct_relevel()`, `fct_lump()` |
| `{purrr}` | Iteration without explicit loops | `map()`, `map_dbl()`, `map2()` |

::: callout-tip
For intermediate users: once your pipelines feel repetitive, wrap recurring logic in a function and use `purrr::map()` over many files or models.
:::

## `ggplot2` teaser (full lesson later)

```{r}
penguins |>
  filter(!is.na(flipper_length_mm), !is.na(body_mass_g), !is.na(species)) |>
  ggplot(aes(flipper_length_mm, body_mass_g, color = species)) +
  geom_point(alpha = 0.75, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 1) +
  scale_color_brewer(palette = "Dark2") +
  labs(
    title = "Bigger flippers, heavier penguins",
    subtitle = "A quick taste of what ggplot2 can do",
    x = "Flipper length (mm)",
    y = "Body mass (g)",
    color = "Species"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "top")
```

# Further reading

[Tidyverse site](https://www.tidyverse.org/){.external target="_blank"}

[dplyr overview](https://dplyr.tidyverse.org/){.external target="_blank"}

[tidyr pivot vignette](https://tidyr.tidyverse.org/articles/pivot.html){.external target="_blank"}
