---
title: "More Tidyverse"
---

```{r}
#| output: false

library(tidyverse)
set.seed(1234)
```

The **tidyverse** is a collection of packages built for a common style of data analysis.
The idea is simple: keep data in tidy columns, then solve tasks by chaining clear verbs.

## Some data to work on

```{r}
n <- 1000
df <- tibble(
  id = 1:n,
  date = sample(
    seq(ymd("2024-01-01"), ymd("2024-12-31"), by = "days"), 
    n, replace = TRUE
  ),
  age = rnorm(n, mean = 40, sd = 10) |> round(),
  sex = sample(
    c("M", "F"), 
    n, replace = TRUE
  ),
  weight = rnorm(n, mean = 75, sd = 10) |> round(),
  height = (rnorm(n, mean = 163, sd = 7) + (date - ymd("2024-01-01")) / 36.5 + (sex == "M")*7) |> as.numeric() |> round(),
  note = sample(
    c("new-patient", "follow-up", "urgent"), 
    n, replace = TRUE
  ) |> factor()
)

df |> head(10) |> gt::gt(id = "patients_tbl")
```

# `dplyr` in one minute

`{dplyr}` is for data manipulation with readable verbs.
Most common pattern: `data |> filter(...) |> mutate(...) |> summarise(...)`.

# `tidyr`: reshape data

`{tidyr}` helps move between **wide** and **long** data and fill missing values.

## `pivot_longer()` and `pivot_wider()`

These are ways of reshaping data. Let's start with a dataset in wide format:
```{r}
mortality_df <- tibble(
  group = c("A", "B"),
  mort_6mo = c(0.1,0.15),
  mort_1yr = c(0.2,0.25),
  mort_2yr = c(0.3,0.35),
  mort_5yr = c(0.5,0.55)
)
mortality_df |> gt::gt()
```

Now let's make it long:
```{r}
mortality_df_long <- mortality_df |> pivot_longer(
  cols = c(mort_6mo, mort_1yr, mort_2yr, mort_5yr),
  names_to = "time",
  values_to = "mortality",
  names_pattern = "mort_(.*)"
)

mortality_df_long |> gt::gt()
```

To go the other way, we can widen it back (notice, we have control over how column names are transformed into row values, and vice versa):
```{r}
mortality_df_long |> pivot_wider(
  names_from = time,
  names_glue = "{time}_{.value}",
  values_from = mortality
)  |> gt::gt()

```

## `separate_wider_delim()`

A simple way of splitting a column into multiple columns based on a delimiter:
```{r}
data.frame(
  x = c("A_1", "B_2", "C_1", "D_2", "E_2")
) |> gt::gt()

data.frame(
  x = c("A_1", "B_2", "C_1", "D_2", "E_2")
) |> separate_wider_delim(
  x, delim = "_",
  names = c("letter", "number")
) |> gt::gt()
```

# Other tidyverse packages (quick tour)

## `readr` (import text data)

Purpose: fast, friendly file import.
Useful functions: `read_csv()`, `read_tsv()`, `write_csv()`.

::: callout-tip 
Other nice packages are `{readxl}` for Excel files and `{googlesheets4}` for Google Sheets along with `{tidyxl}` for non-tabular Excel files.
:::

## `tibble` (modern data frame)

Tibbles are a special type of `data.frame`, providing a few benefits; cleaner printing and safer column handling. In addition, building tibbles allow you to build the columns sequentially (i.e., you can refer to columns created earlier in the same `tibble()` call).
```{r}
tibble(
  x = 1:3,
  y = x*3
) |> gt::gt()
```


Also comes with `add_row()`:
```{r}
df |> 
  head(5) |>
  add_row(id = -1, date = ymd("2024-05-05"), sex = "M") |> 
  gt::gt()
```

## `purrr` (iterate without explicit loops)

A clean way of applying functions to subsets (see <a href="05-Loops.html">previous lesson on Loops and parallelisation</a>).
Useful functions: `map()`, `map_dbl()`, `map2()`.

```{r}
df |> 
  select(weight, sex) |>
  group_by(sex) |>
  summarise(mean_weight = mean(weight)) |>
  gt::gt()

split(df$weight, df$sex) |> 
  purrr::map_dfr(mean) |> gt::gt()
```

## `stringr` (string handling)

Purpose: consistent string tools.
Useful functions: `str_detect()`, `str_replace()`, `str_extract()`.

```{r}
c("word", "hypenated-word", "unhyphenated word") |> str_detect("-")

df |> head(5) |> gt::gt()

df |> 
  filter(
    str_detect(note, "-")
  ) |> 
  mutate(
    note_clean = str_replace(note, "-", " ")
  ) |> 
  head(5) |> gt::gt()
```

## `forcats` (factor handling)

Purpose: robust factor workflows.
Useful functions: `fct_relevel()`, `fct_reorder()`, `fct_lump()`.

```{r}
df$note |> levels()

z <- df |> mutate(
  note = fct_relevel(note, "follow-up", after = 3)
)
z$note |> levels()
```

## `lubridate` (dates and times)

Purpose: parse and manipulate dates.
Useful functions: `ymd()`, `wday()`, `floor_date()`.

```{r}
df |>
  select(id, date) |>
  mutate(
    weekday = wday(date, label = T),
    prior_monday = floor_date(date, "week", week_start = 1)
  ) |> 
  head(5) |> gt::gt()
```

# `ggplot2`: a future lesson topic

```{r}
library(wesanderson)
df |>
  ggplot(aes(x = date, y = height, color = sex, fill = sex)) +
  geom_point(size = 1.5, alpha = 0.5, shape =20) +
  geom_smooth(method = "loess", formula = y~x, linewidth = 1, alpha = 0.5) +
  labs(
    title = "A plot of height over time",
    subtitle = "The grammar of graphics—building plots by adding layers",
    x = "Date",
    y = "Height (cm)",
    color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    legend.position = "top",
    axis.title.x = element_text(margin = margin(t = 20)),
    axis.title.y = element_text(margin = margin(r = 20))
  ) +
  scale_color_manual(values = wes_palette("Royal1"), labels = c("F" = "Female", "M" = "Male")) +
  scale_fill_manual(values = wes_palette("Royal1"), labels = c("F" = "Female", "M" = "Male"), name=NULL)
```

# Further reading

[R for Data Science: Data transform](https://r4ds.hadley.nz/data-transform.html){.external target="_blank"}

[R for Data Science: Tidy data](https://r4ds.hadley.nz/data-tidy.html){.external target="_blank"}

[Tidyverse documentation](https://www.tidyverse.org/packages/){.external target="_blank"}
