---
title: "More Tidyverse"
---

```{r}
#| output: false

library(tidyverse)
library(lubridate)
set.seed(1234)
```

The **tidyverse** is a collection of packages built for a common style of data analysis.
The idea is simple: keep data in tidy columns, then solve tasks by chaining clear verbs.

## Some data to work on

```{r}
patients <- tibble(
  id = 1:8,
  date = c("2026-04-01", "2026-04-02", "2026-04-02", "2026-04-03", "2026-04-03", "2026-04-04", "2026-04-04", "2026-04-05"),
  sex = c("Female", "Male", "Female", "Male", "Female", "Female", "Male", "Female"),
  ward = c("A", "A", "B", "B", "A", "B", "A", "B"),
  systolic = c(128, 142, 135, 150, 131, 138, 145, 126),
  note = c("follow-up in 2 weeks", "new referral", "repeat blood test", "urgent review", "exercise started", "diet plan", "new medication", "stable")
) |>
  mutate(date = ymd(date))

patients |> gt::gt(id = "patients_tbl")
```

# `dplyr` in one minute

`{dplyr}` is for data manipulation with readable verbs.
Most common pattern: `data |> filter(...) |> mutate(...) |> summarise(...)`.

```{r}
patients |>
  filter(systolic >= 140) |>
  select(id, sex, ward, systolic)
```

# `tidyr`: reshape data

`{tidyr}` helps move between **wide** and **long** data and fill missing values.

## `pivot_longer()` and `pivot_wider()`

```{r}
vitals_wide <- patients |>
  transmute(id, systolic, diastolic = round(systolic * 0.65))

vitals_long <- vitals_wide |>
  pivot_longer(cols = c(systolic, diastolic), names_to = "measure", values_to = "value")

vitals_long

vitals_long |>
  pivot_wider(names_from = measure, values_from = value)
```

## `separate_wider_delim()`

```{r}
patients |>
  mutate(tag = paste(ward, sex, sep = "_")) |>
  separate_wider_delim(tag, delim = "_", names = c("ward2", "sex2")) |>
  select(id, ward2, sex2)
```

# Other tidyverse packages (quick tour)

## `readr` (import text data)

Purpose: fast, friendly file import.
Useful functions: `read_csv()`, `read_tsv()`, `write_csv()`.

```{r}
#| eval: false
raw <- readr::read_csv("patients.csv")
readr::write_csv(raw, "patients-clean.csv")
```

## `tibble` (modern data frame)

Purpose: cleaner printing and safer column handling.
Useful functions: `tibble()`, `add_row()`, `as_tibble()`.

```{r}
patients |>
  add_row(id = 9, date = ymd("2026-04-06"), sex = "Male", ward = "A", systolic = 133, note = "phone follow-up") |>
  tail(2)
```

## `purrr` (iterate without explicit loops)

Purpose: apply functions consistently.
Useful functions: `map()`, `map_dbl()`, `map2()`.

```{r}
patients |>
  select(systolic) |>
  summarise(across(everything(), list(mean = mean, sd = sd)))

split(patients$systolic, patients$ward) |>
  purrr::map_dbl(mean)
```

## `stringr` (string handling)

Purpose: consistent string tools.
Useful functions: `str_detect()`, `str_replace()`, `str_extract()`.

```{r}
patients |>
  filter(str_detect(note, "new|urgent")) |>
  mutate(note_clean = str_replace(note, "-", " ")) |>
  select(id, note, note_clean)
```

## `forcats` (factor handling)

Purpose: robust factor workflows.
Useful functions: `fct_relevel()`, `fct_reorder()`, `fct_lump()`.

```{r}
patients |>
  mutate(
    ward = fct_relevel(ward, "B", "A"),
    sex = fct_lump(factor(sex), n = 1)
  ) |>
  count(ward, sex)
```

## `lubridate` (dates and times)

Purpose: parse and manipulate dates.
Useful functions: `ymd()`, `wday()`, `floor_date()`.

```{r}
patients |>
  mutate(
    weekday = wday(date, label = TRUE),
    week_start = floor_date(date, unit = "week")
  ) |>
  select(id, date, weekday, week_start)
```

# `ggplot2` teaser (full lesson later)

One line summary: `{ggplot2}` builds plots by adding layers.

```{r}
patients |>
  ggplot(aes(x = date, y = systolic, color = sex)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Blood pressure over time",
    subtitle = "Lesson 7 teaser: grammar of graphics",
    x = NULL,
    y = "Systolic blood pressure (mmHg)",
    color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

# Further reading

[R for Data Science: Data transform](https://r4ds.hadley.nz/data-transform.html){.external target="_blank"}

[R for Data Science: Tidy data](https://r4ds.hadley.nz/data-tidy.html){.external target="_blank"}

[Tidyverse documentation](https://www.tidyverse.org/packages/){.external target="_blank"}
