---
title: "More Tidyverse"
---

```{r}
#| output: false

library(tidyverse)
library(lubridate)
set.seed(1234)
```

The **tidyverse** is a collection of packages built for a common style of data analysis.
The idea is simple: keep data in tidy columns, then solve tasks by chaining clear verbs.

## Some data to work on

```{r}
n <- 1000
df <- tibble(
  id = 1:n,
  date = sample(seq(ymd("2024-01-01"), ymd("2024-12-31"), by = "days"), n, replace = TRUE),
  age = rnorm(n, mean = 40, sd = 10) |> round(),
  sex = sample(c("M", "F"), n, replace = TRUE),
  weight = rnorm(n, mean = 75, sd = 10) |> round(),
  height = (rnorm(n, mean = 170, sd = 7) + (date - ymd("2024-01-01")) / 36.5 - (sex == "F")*7) |> as.numeric() |>round(),
  note = sample(c("new-patient", "follow-up", "urgent"), n, replace = TRUE)
)

df |> head(10) |> gt::gt(id = "patients_tbl")
```

# `dplyr` in one minute

`{dplyr}` is for data manipulation with readable verbs.
Most common pattern: `data |> filter(...) |> mutate(...) |> summarise(...)`.

# `tidyr`: reshape data

`{tidyr}` helps move between **wide** and **long** data and fill missing values.

## `pivot_longer()` and `pivot_wider()`

```{r}
df_wide <- df |>
  select(id, weight, height)

df_long <- df_wide |>
  pivot_longer(cols = c(weight, height), names_to = "measure", values_to = "value")

df_long

df_long |>
  pivot_wider(names_from = measure, values_from = value)
```

## `separate_wider_delim()`

```{r}
data.frame(
  x = c("A_1", "B_2", "C_1", "D_2", "E_2")
) |>
  separate_wider_delim(x, delim = "_", names = c("letter", "number"))
```

# Other tidyverse packages (quick tour)

## `readr` (import text data)

Purpose: fast, friendly file import.
Useful functions: `read_csv()`, `read_tsv()`, `write_csv()`.

::: callout-tip 
Other nice packages are `{readxl}` for Excel files and `{googlesheets4}` for Google Sheets along with `{tidyxl}` for non-tabular Excel files.
:::

## `tibble` (modern data frame)

Purpose: cleaner printing and safer column handling.
Useful functions: `tibble()`, `add_row()`, `as_tibble()`.

```{r}
df |>
  add_row(id = 9, date = ymd("2026-04-06"), sex = "Male")
```

## `purrr` (iterate without explicit loops)

Purpose: apply functions consistently.
Useful functions: `map()`, `map_dbl()`, `map2()`.

```{r}
df |>
  select(weight, height) |>
  summarise(across(everything(), list(mean = mean, sd = sd)))

split(df$weight, df$sex) |>
  purrr::map_dbl(mean)
```

## `stringr` (string handling)

Purpose: consistent string tools.
Useful functions: `str_detect()`, `str_replace()`, `str_extract()`.

```{r}
df |>
  filter(str_detect(note, "new|urgent")) |>
  mutate(note_clean = str_replace(note, "-", " ")) |>
  select(id, note, note_clean)
```

## `forcats` (factor handling)

Purpose: robust factor workflows.
Useful functions: `fct_relevel()`, `fct_reorder()`, `fct_lump()`.

```{r}
df |>
  mutate(
    sex = fct_lump(factor(sex), n = 1)
  ) |>
  count(sex)
```

## `lubridate` (dates and times)

Purpose: parse and manipulate dates.
Useful functions: `ymd()`, `wday()`, `floor_date()`.

```{r}
df |>
  mutate(
    weekday = wday(date, label = TRUE),
    week_start = floor_date(date, unit = "week")
  ) |>
  select(id, date, weekday, week_start)
```

# `ggplot2` teaser (full lesson later)

One line summary: `{ggplot2}` builds plots by adding layers.

```{r}
df |>
  ggplot(aes(x = date, y = height, color = sex)) +
  geom_point(size = 3, alpha = 0.9) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1) +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Height over time",
    subtitle = "Lesson 7 teaser: grammar of graphics",
    x = NULL,
    y = "Height (cm)",
    color = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "top")
```

# Further reading

[R for Data Science: Data transform](https://r4ds.hadley.nz/data-transform.html){.external target="_blank"}

[R for Data Science: Tidy data](https://r4ds.hadley.nz/data-tidy.html){.external target="_blank"}

[Tidyverse documentation](https://www.tidyverse.org/packages/){.external target="_blank"}
