---
title: "Working with many packages"
---

```{r}
#| output: false

library(dplyr)
library(ggplot2)
library(broom)
library(gtsummary)
library(conflicted)
set.seed(1234)
```

In practice, most R analyses load a handful of packages at once.
That's great — each package solves a specific problem — but it also creates two challenges:

1. **Conflicts**: two packages may export a function with the same name.
2. **Compatibility**: does the output of one package feed naturally into the next?

This lesson covers both.

## Some data to work on

```{r}
n <- 500
df <- tibble(
  id      = 1:n,
  age     = rnorm(n, mean = 55, sd = 12) |> round(),
  sex     = sample(c("Male", "Female"), n, replace = TRUE),
  smoker  = sample(c(TRUE, FALSE), n, replace = TRUE, prob = c(0.3, 0.7)),
  sbp     = round(120 + 0.4 * age + 8 * smoker + rnorm(n, sd = 12)),
  bmi     = round(rnorm(n, mean = 26, sd = 4), 1)
)

df |> head(8) |> gt::gt(id = "patient_tbl")
```

# Loading order and function masking

Every time you call `library(pkg)`, R **attaches** the package to the *search path* — a list of environments R looks through, left to right, when you type a name.

```{r}
#| eval: false
search()
```

```
 [1] ".GlobalEnv"       "package:conflicted" "package:gtsummary"
 [4] "package:broom"    "package:ggplot2"    "package:dplyr"
 [7] "package:stats"    "package:base"
```

The last package you load sits at position 2 (just after `.GlobalEnv`), so its functions are found first.
When two packages share a function name, the one loaded *later* silently **masks** the earlier one.

## A concrete example

Both `{dplyr}` and base `{stats}` have a function called `filter()`.
After `library(dplyr)`, typing `filter()` calls the dplyr version — the stats one is still there, just hidden.

```{r}
#| eval: false
library(MASS)    # also exports select()
library(dplyr)   # also exports select()
#> Attaching package: 'dplyr'
#> The following object is masked from 'package:MASS':
#>   select
```

R warns you, but only if you read the message.
In a busy session these messages are easy to miss.

## Three solutions to masking

### 1. Explicit namespacing with `::`

The safest habit: always write `pkg::function()` for any function you care about.

```{r}
# Always unambiguous:
df |> dplyr::select(id, age, sbp) |> head(3)
```

This works even when a function is masked, and makes code easier to read and share.

### 2. The `{conflicted}` package

`{conflicted}` turns silent masking into a loud error, forcing you to be explicit.
Load it at the top of your script and every ambiguous call will stop with a helpful message.

```{r}
#| error: true

# conflicted is already loaded above; now any ambiguous call errors:
filter(df, age > 70)
```

You resolve the conflict once using `conflict_prefer()`:

```{r}
conflict_prefer("filter", "dplyr")
```

Now `filter()` unambiguously refers to the dplyr version for the rest of the session.

```{r}
df |> filter(age > 70) |> dplyr::select(id, age, sbp) |> head(5)
```

### 3. Load packages in a deliberate order

If you don't want to use `{conflicted}`, load packages in order of priority — the last one loaded wins.
However, this is the least robust approach and is not recommended in shared code.

::: callout-tip
**Best practice**: use `::` for any function that might be ambiguous, and load `{conflicted}` in
scripts you share with others. You get both safety and clarity.
:::

# Cross-package workflows

A lot of R's power comes from packages that are designed to **talk to each other**.
The key idea: if a function always returns a standard, predictable object (like a data frame or a specific class), other packages can be written to work with it directly.

## Generic functions and methods

Base R uses **generics** — functions like `summary()`, `print()`, and `predict()` — that behave differently depending on what they are given.

```{r}
m <- lm(sbp ~ age + sex + smoker + bmi, data = df)

# summary() "knows" it has a linear model and formats accordingly:
summary(m)
```

This system — called **S3 dispatch** — means that different packages can teach the same verb new tricks, without you having to learn a different function name.

## Standardised outputs: the tidy principle

The tidyverse is built on a simple idea: **data in, data out**.
Each function takes a data frame and returns a data frame, making operations easy to chain.

Many packages outside the core tidyverse follow the same principle.
The key package here is `{broom}`, which converts messy model objects into tidy tibbles.

```{r}
# A linear model gives an opaque object:
class(m)
m   # hard to work with programmatically
```

```{r}
# broom::tidy() extracts the coefficients as a clean tibble:
broom::tidy(m, conf.int = TRUE)
```

```{r}
# broom::glance() gives one-row model-level summaries:
broom::glance(m)
```

```{r}
# broom::augment() adds fitted values and residuals back to your data:
broom::augment(m, df) |> 
  dplyr::select(id, sbp, .fitted, .resid) |> 
  head(5)
```

Because `tidy()` always returns a tibble with the same column names (`term`, `estimate`, `std.error`, `p.value`, …), the output plugs directly into `{ggplot2}`:

```{r}
broom::tidy(m, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = term)) +
  geom_pointrange() +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    title = "Linear model coefficients",
    x = "Estimate (95 % CI)",
    y = NULL
  ) +
  theme_minimal(base_size = 13)
```

## `{gtsummary}`: a complete cross-package workflow

`{gtsummary}` is a good example of a package that slots into a tidy workflow seamlessly.
It takes standard R model objects and produces publication-ready tables, using the same underlying ideas.

```{r}
# Descriptive table — takes a data frame, returns a gt table:
df |>
  dplyr::select(age, sex, smoker, bmi, sbp) |>
  tbl_summary(
    by = smoker,
    label = list(
      age   ~ "Age (years)",
      sex   ~ "Sex",
      bmi   ~ "BMI (kg/m²)",
      sbp   ~ "Systolic BP (mmHg)"
    )
  ) |>
  add_p() |>
  add_overall() |>
  bold_labels()
```

```{r}
# Regression table — takes any model object directly:
m |>
  tbl_regression(
    label = list(
      age    ~ "Age (years)",
      sex    ~ "Sex",
      smoker ~ "Smoker",
      bmi    ~ "BMI (kg/m²)"
    )
  ) |>
  bold_p() |>
  bold_labels()
```

No need to manually extract and format coefficients — `{gtsummary}` calls `broom::tidy()` internally, giving you a formatted table in two lines.

::: callout-note
`{gtsummary}` works with a wide range of model types: `lm`, `glm`, `coxph` (survival), `lme4` mixed models, and more — all using the same `tbl_regression()` call.
:::

## Another example: survival analysis

The same pattern works with survival models from the `{survival}` package:

```{r}
library(survival)

# Simulate a time-to-event outcome:
df2 <- df |> mutate(
  time  = rexp(n, rate = 0.05),
  event = rbinom(n, 1, prob = 0.6)
)

cox_m <- coxph(Surv(time, event) ~ age + sex + smoker, data = df2)

# broom and gtsummary work exactly the same way:
broom::tidy(cox_m, exponentiate = TRUE, conf.int = TRUE)
```

```{r}
cox_m |>
  tbl_regression(
    exponentiate = TRUE,
    label = list(
      age    ~ "Age (years)",
      sex    ~ "Sex",
      smoker ~ "Smoker"
    )
  ) |>
  bold_p() |>
  bold_labels()
```

# `{box}`: surgical package imports

An alternative to `library()` is the `{box}` package.
Instead of loading an entire package namespace, you import only the specific functions you need.

```{r}
#| eval: false

box::use(
  dplyr[filter, select, mutate],   # only these three functions from dplyr
  ggplot2[ggplot, aes, geom_point] # only these from ggplot2
)

# Nothing else from dplyr or ggplot2 is in scope — no masking possible
```

This is especially useful in packages, larger codebases, or when you want the code to be completely explicit about where every function comes from.

::: callout-tip
For day-to-day scripts, `library()` + `{conflicted}` is usually enough.
`{box}` becomes attractive in package development or when you want strict, reproducible namespacing.
:::

# Key takeaways

| Problem | Solution |
|---|---|
| Silent function masking | Use `pkg::fn()` or load `{conflicted}` |
| Want to resolve a conflict once | `conflict_prefer("fn", "pkg")` |
| Model output hard to work with | `broom::tidy()` / `broom::glance()` |
| Need a publication table fast | `gtsummary::tbl_regression()` |
| Want zero namespace pollution | `box::use(pkg[fn1, fn2])` |

# Further reading

[`{conflicted}` package documentation](https://conflicted.r-lib.org/){.external target="_blank"}

[`{broom}` package documentation](https://broom.tidymodels.org/){.external target="_blank"}

[`{gtsummary}` package documentation](https://www.danieldsjoberg.com/gtsummary/){.external target="_blank"}

[R for Data Science – Functions (namespacing)](https://r4ds.hadley.nz/functions.html){.external target="_blank"}

[`{box}` package documentation](https://klmr.me/box/){.external target="_blank"}
