---
title: "Working with many packages"
---

```{r}
#| output: false

library(dplyr)
library(ggplot2)
library(broom)
library(gtsummary)
set.seed(1234)
```

In practice, most R analyses load a handful of packages at once.
That's great — each package solves a specific problem — but it also creates two challenges:

1. **Conflicts**: two packages may export a function with the same name.
2. **Compatibility**: does the output of one package feed naturally into the next?

This lesson covers both.

## Some data to work on

```{r}
n <- 500
df <- tibble(
  id      = 1:n,
  age     = rnorm(n, mean = 55, sd = 12) |> round(),
  sex     = sample(c("Male", "Female"), n, replace = TRUE),
  smoker  = sample(c(TRUE, FALSE), n, replace = TRUE, prob = c(0.3, 0.7)),
  sbp     = round(120 + 0.4 * age + 8 * smoker + rnorm(n, sd = 12)),
  bmi     = round(rnorm(n, mean = 26, sd = 4), 1)
)

df |> head(8) |> gt::gt(id = "patient_tbl")
```

# Loading order and function masking

Every time you call `library(pkg)`, R **attaches** the package to the *search path* — a list of environments R looks through, left to right, when you type a name.

```{r}
search()
```

The last package you load sits at position 2 (just after `.GlobalEnv`), so its functions are found first.
When two packages share a function name, the one loaded *later* silently **masks** the earlier one.

## A concrete example

Both `{dplyr}` and `{MASS}` have a function called `select()`.
After `library(MASS)`, typing `select()` calls the MASS version — the dplyr one is still there, just hidden.

```{r}
library(dplyr)
library(MASS)
```

R warns you but proceeds - there's no guarantee you will notice the warning message in a busy session.

## Four solutions to masking

### 1. Explicit namespacing with `::`

The safest habit: write `pkg::function()` for any function that might be ambiguous.

```{r}
# Always unambiguous:
df |> dplyr::select(id, age, sbp) |> head(3)
```

This works even when a function is masked, and makes code easier to read and share.

### 2. The `{conflicted}` package

`{conflicted}` turns silent masking into a loud error, forcing you to be explicit.
Load it at the top of your script and every ambiguous call will stop with a helpful message.

```{r}
#| error: true

library(conflicted)

# conflicted is already loaded above; now any ambiguous call errors:
filter(df, age > 70)
```

You resolve the conflict once using `conflict_prefer()`:

```{r}
conflict_prefer("filter", "dplyr")
```

Now `filter()` unambiguously refers to the dplyr version for the rest of the session.

```{r}
df |> filter(age > 70) |> dplyr::select(id, age, sbp) |> head(5)
```

### 3. Load packages in a deliberate order

If you don't want to use `{conflicted}`, load packages in order of priority — the last one loaded wins.
Note, this is the least robust approach.

### 4. `{box}`: surgical package imports

An alternative to `library()` is the `{box}` package.
Instead of loading an entire package namespace, you import only the specific functions you need.

```{r}
#| eval: false

box::use(
  dplyr[filter, select, mutate],   # only these three functions from dplyr
  ggplot2[ggplot, aes, geom_point], # only these from ggplot2
  MASS(mass_filter = filter) # rename MASS::filter as mass_filter to avoid conflict with dplyr
)

# Nothing else from dplyr or ggplot2 is in scope — no masking possible
```

This is especially useful in packages, larger codebases, or when you want the code to be completely explicit about where every function comes from.

::: callout-tip
For day-to-day scripts, `library()` with explicit namespacing (`pkg::function()`) + `{conflicted}` is usually enough.

`{box}` becomes attractive in package development or when you want strict, reproducible namespacing.
:::

# Cross-package workflows

A lot of R's power comes from packages that are designed to **talk to each other**.
The key idea: if a function always returns a standard, predictable object (like a data frame or a specific class), other packages can be written to work with it directly.

## Generic functions and methods

Base R uses **generics** — functions like `summary()`, `print()`, and `predict()` — that behave differently depending on what they are given.

```{r}
m <- lm(sbp ~ age + sex + smoker + bmi, data = df)

# summary() "knows" it has a linear model and formats accordingly:

summary(m)
```

This system — called **S3 dispatch** — means that different packages can teach the same verb new tricks, without you having to learn a different function name.

Under the hood, R looks at the class of the object you pass to `summary()`, sees it's a `lm` object, and calls `summary.lm()` automatically.

## Standardised outputs: the tidy principle

The tidyverse is built on a simple idea: **data in, data out**.
Each function takes a data frame and returns a data frame, making operations easy to chain.

Many packages outside the core tidyverse follow the same principle.
The key package here is `{broom}`, which converts messy model objects into tidy tibbles.

```{r}
# A linear model gives an opaque object:
class(m)
m   # hard to work with programmatically
```

```{r}
# broom::tidy() extracts the coefficients as a clean tibble:
broom::tidy(m, conf.int = TRUE)
```

```{r}
# broom::glance() gives one-row model-level summaries:
broom::glance(m)
```

```{r}
# broom::augment() adds fitted values and residuals back to your data:
broom::augment(m, df) |> 
  dplyr::select(id, sbp, .fitted, .resid) |> 
  head(5)
```

Because `tidy()` always returns a tibble with the same column names (`term`, `estimate`, `std.error`, `p.value`, …), the output easily plugs into `{ggplot2}`:

```{r}
broom::tidy(m, conf.int = TRUE) |>
  filter(term != "(Intercept)") |>
  ggplot(aes(x = estimate, xmin = conf.low, xmax = conf.high, y = term)) +
  geom_pointrange()
```

## `{gtsummary}`: a complete cross-package workflow

`{gtsummary}` is a good example of a package that slots into a tidy workflow seamlessly.
It takes standard R model objects and produces publication-ready tables, using the same underlying ideas.

```{r}
# Regression table — takes any model object directly:
m |>
  tbl_regression(
    label = list(
      age    ~ "Age (years)",
      sex    ~ "Sex",
      smoker ~ "Smoker",
      bmi    ~ "BMI (kg/m²)"
    )
  ) |>
  bold_p() |>
  bold_labels()
```

No need to manually extract and format coefficients — `{gtsummary}` calls `broom::tidy()` internally, giving you a formatted table in two lines.

::: callout-note
`{gtsummary}` works with a wide range of model types: `lm`, `glm`, `coxph` (survival), `lme4` mixed models, and more — all using the same `tbl_regression()` call.
:::

## Another example: survival analysis

The same pattern works with survival models from the `{survival}` package:

```{r}
library(survival)

# Simulate a time-to-event outcome:
df2 <- df |> mutate(
  time  = rexp(n, rate = 0.05),
  event = rbinom(n, 1, prob = 0.6)
)

cox_m <- coxph(Surv(time, event) ~ age + sex + smoker, data = df2)
```

`{broom}` and `{gtsummary}` work exactly the same way:
```{r}
cox_m |> broom::tidy(exponentiate = TRUE, conf.int = TRUE)

cox_m |>
  tbl_regression(
    exponentiate = TRUE)
```

# A more general principle: design for interoperability

## The tidy principle
Functions that take and return data frames with predictable structure — is a powerful design pattern that encourages interoperability between packages.

When you choose packages for your workflow, look for ones that follow this principle.

## Understanding function inputs and outputs
When you use a new package, take time to understand what its functions expect as input and what they return.

Start with `?pkg::fn` to read the documentation and go to the package website / vignettes (= *user manuals*). Then experiment with simple examples to see the output structure.

Once you know the input requirements and the output format, try playing with the output yourself. Take you time to manually inspect and manipulate the output into different formats; e.g., tables, plots, etc.

Once you understand this, it's time to see how you can use the output in your workflow, and which other packages fit well into the workflow.

# Key takeaways

| Problem | Solution |
|---|---|
| Silent function masking | Use `pkg::fn()` or load `{conflicted}` |
| Want to resolve a conflict once | `conflict_prefer("fn", "pkg")` |
| Want zero namespace pollution | `box::use(pkg[fn1, fn2])` |
| Understand your function inputs/outputs | `?pkg::fn` |
| Model output hard to work with | `broom::tidy()` / `broom::glance()` |
| Need a publication table fast | `gtsummary::tbl_regression()` |

# Further reading

[`{conflicted}` package documentation](https://conflicted.r-lib.org/){.external target="_blank"}

[`{broom}` package documentation](https://broom.tidymodels.org/){.external target="_blank"}

[`{gtsummary}` package documentation](https://www.danieldsjoberg.com/gtsummary/){.external target="_blank"}

[R for Data Science – Functions (namespacing)](https://r4ds.hadley.nz/functions.html){.external target="_blank"}

[`{box}` package documentation](https://klmr.me/box/){.external target="_blank"}
