---
title: "Example pipeline"
---

## Working with data from Statistics Denmark (DST)

Here, we'll use the simulated data from [the previous note](https://r-coding-cafe.zheer.dk/notes/extra/0-simulate-data.html){.external target="_blank"} to show how a full data preprocessing pipeline might look. In practice, you would load in the data from DST using the appropriate functions (e.g. read_dta, read_csv, etc.) and then follow a similar pipeline to clean and prepare the data for analysis.


```{r}
#| output: false
library(dplyr)
```

Loading in the simulated data here, but in practice you would load in the data from DST using the appropriate functions (e.g. read_dta, read_csv, etc.):
```{r}
load("simulated_data.RData")
```

### Identify first hip fracture
```{r}
lpr <- left_join(lpr_adm, lpr_diag, by = "RECNUM") |>
  # Only care about A-diagnoses pertaining to hip fracture (S72...)
  filter(C_DIAGTYPE == "A",
         grepl("^S72", C_DIAG)
  ) |>
  # For each person, keep only first date
  group_by(PNR) |>
  filter(D_INDDTO == min(D_INDDTO)) |>
  # Keep one row if multiple diagnoses on the same date
  slice(1) |>
  ungroup() # Always best to ungroup in the end
```

Preview:
```{r}
lpr |> head(3) |> gt::gt()
```

### Add background info
```{r}
pop <- left_join(lpr, bef, by = "PNR") |>
  # calculate age in days, divide into years, round to two decimals
  mutate(age = ((D_INDDTO - FOED_DAG) / 365.25) |> as.numeric() |> round(2)) |>
  select(c(PNR, D_INDDTO, FOED_DAG, koen, age)) |>
  rename(index_date = D_INDDTO)
```

Preview:
```{r}
pop |> head(3) |> gt::gt()
```

### Get last eGFR before fracture
```{r}
last_eGFR <- left_join(pop, lab_dm_forsker, by = "PNR")
# Note, approx 10% of individuals have never had an eGFR measured; they now have NAs

last_eGFR <- last_eGFR |>
  # Get eGFRs prior to fracture
  filter(SAMPLINGDATE <= index_date) |>
  # keep only last eGFR for each person
  group_by(PNR) |>
  filter(SAMPLINGDATE == max(SAMPLINGDATE)) |>
  slice(1) |>
  ungroup() |>
  # keep only necessary columns to add back onto the full pop
  select(c(PNR, SAMPLINGDATE, VALUE)) |>
  rename(eGFR_val = VALUE, eGFR_date = SAMPLINGDATE)

pop <- left_join(pop, last_eGFR, by ="PNR")
```

Previews:
```{r}
last_eGFR |> head(3) |> gt::gt()
pop |> head(3) |> gt::gt()
```

### Get first glucose-lowering drug prescription
```{r}
lmdb_first_gld <- lmdb |>
  # only get glucose-lowering drugs (ATC code starting with A10BA02, A10BB12, A10BH0[1,2,3], A10BJ0[2,6], or A10BK0[1-4])
  filter(grepl("^A10BA02|^A10BB12|^A10BH0[1,2,3]|^A10BJ0[2,6]|^A10BK0[1-4]", ATC)) |>
  # For each person, only keep the earliest date of redemption
  group_by(PNR) |>
  filter(EKSD == min(EKSD)) |>
  # keep one line per person
  slice(1) |>
  ungroup() |>
  select(c(PNR, EKSD))
```

Preview:
```{r}
lmdb_first_gld |> head(3) |> gt::gt()
```

### Add in medication data and remove individuals with prior drug exposure
```{r}
pop <- left_join(pop, lmdb_first_gld, by ="PNR")

# Only keep people with no prior glucose-lowering drug use before hip fracture
pop <- pop |>
  filter(is.na(EKSD) | index_date <= EKSD) |>
  select(-c(EKSD))
```

Preview:
```{r}
pop |> head(3) |> gt::gt()
```

### Get first metformin prescription
```{r}
lmdb_metformin <- lmdb |>
  filter(ATC == "A10BA02")

lmdb_metformin <- lmdb_metformin |>
  # For each person, only keep the earliest date of redemption
  group_by(PNR) |>
  filter(EKSD == min(EKSD)) |>
  # For each prescription, multiply the pill dosage by the number of pills to get a total dose and sum it over all prescriptions for that date (in case of multiple redemptions same day) to get a grand total
  mutate(no_of_pills = sum(APK * PACKSIZE),
         total_dose = sum(APK * PACKSIZE * STRNUM)) |>
  # keep one line per person
  slice(1) |>
  ungroup() |>
  select(c(PNR, EKSD, no_of_pills, total_dose)) |>
  rename(metf_date = EKSD,
         metf_no_of_pills = no_of_pills,
         metf_total_dose = total_dose)
```

Preview:
```{r}
lmdb_metformin |> head(3) |> gt::gt()
```

### Add in metformin data
```{r}
pop <- left_join(pop, lmdb_metformin, by ="PNR")
```

Preview:
```{r}
pop |> head(3) |> gt::gt()
```

### Calculate time to metformin
```{r}
pop <- pop |>
  mutate(tte_metf = difftime(metf_date, index_date))
```

Preview:
```{r}        
pop |> head(3) |> gt::gt()
```

## Now you can start analysing the data...
```{r}
library(ggplot2)
ggplot(pop, aes(x=age, y = eGFR_val)) +
  geom_point() +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 3)) +
  theme_bw()
```

## Or save the cleaned dataset for later use
```{r}
saveRDS(pop, "cleaned_data.rds")
```

Making it easy to load the dataset in at a later time without having to repeat the whole pipeline:
```{r}
pop <- readRDS("cleaned_data.rds")
```
